---
title: "Covid"
author: "Klaus Watschinger"
date: "11/25/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("stringr")
library("tidyverse")
library("lubridate")
library("maps")
```

## Load Data
For this report we use data from the Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University. We load for csv files.

```{r get_hoskins_data}
## Get current Data in the four files which are all in the same folder
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <-
  c("time_series_covid19_confirmed_US.csv",
    "time_series_covid19_confirmed_global.csv",
    "time_series_covid19_deaths_US.csv",
    "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)
```

With those 4 urls we now read in the data.

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

## Clean up Data
My initial review of the data leads to the following next steps:

* Tidy data, values for each date should go into a row
* Remove Lat and Long
* Rename Region and State columns
* Convert strings to date
* Join deaths and cases for US and Global data sets.

```{r tidy_global_data, messages = FALSE}
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to ="date", values_to = "cases") %>%
  select(-c(Lat,Long))
global_deaths <- global_deaths %>%
  pivot_longer(cols= -c(`Province/State`, `Country/Region`, Lat, Long), names_to ="date", values_to = "deaths") %>%
  select(-c(Lat,Long))
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))
#take a look at the data after the join, renaming and tidying up
summary(global)
```

Now tidying the US data sets.
```{r tidy_US_Data, messages = FALSE}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))
US <- US_cases %>%
  full_join(US_deaths)
```
```{r modify_Global }
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

We now retrieve ISO and FIPS country codes to complete our work on global data:
```{r ISO_FIPS_Codes, messages = FALSE}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

Finally, we perform a left join (all records in global are matched with the ISO FIPS codes) and finalize our global data set.
```{r join_global_iso}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

```

## Charting
### Preparing the tibbles
Now we are going to visualize the US data. We want to see the data grouped on state level. (per date)
```{r US_by_state, messages = FALSE, warnings = FALSE}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  mutate(cases_per_mill = cases *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, cases_per_mill, Population) %>%
  ungroup()
#Check the latest records
tail(US_by_state)
```
But we also want to see the totals for the US:
```{r US_totals, messages = FALSE}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  mutate(cases_per_mill = cases *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, cases_per_mill, Population) %>%
  ungroup()
#Check the latest records
tail(US_totals)
```

### Timeseries
Now we want to visualize the development of cases and deaths for the US as a whole:
```{r US_Timeseries, messages = FALSE}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10(labels = scales::comma) +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in the US", y= NULL)
```


### Choropleth map
To reflect differences by state, we use a choropleth map based on the latest date available and display the cases per million
```{r Map_states, messages = FALSE}
US_states_latest <- US_by_state %>%
  filter(date == max(date)) %>%
  mutate(Province_State = tolower(Province_State))
##Get map data for US states
MainStates <- map_data("state")
MainStates <- as_tibble(MainStates)
MainStates <- MainStates %>%
  rename(Province_State = region) 
  

US_Map_states_cases <- US_states_latest %>%
  inner_join(MainStates, by = "Province_State") 

US_Map_states_cases %>%
  ggplot(data=US_Map_states_cases, mapping=aes(x=long, y=lat)) +
  geom_polygon(mapping=aes(x=long, y=lat,fill=deaths_per_mill, group=group))

```

## Modeling deaths vs cases on data of the US

### Linear regression
Now we want to develop a linear model to predict deaths depending on observed cases
```{r Linear_Model, messages = FALSE}
US_states_latest <- US_states_latest %>% filter(!(Population == 0))
mod <- lm(deaths_per_mill ~ cases_per_mill, data = US_states_latest)
print(summary(mod))

#US_states_latest %>% slice_max(cases_per_mill)

x_grid <- seq(1, 250000)
new_df <- tibble(cases_per_mill = x_grid)
US_states_pred <- US_states_latest %>% mutate(pred = predict(mod))
US_states_pred %>% ggplot() + geom_point(aes(x = cases_per_mill, y = deaths_per_mill),color="blue") + geom_point(aes(x = cases_per_mill, y = pred), color = "red")
```

### Multivariate regression
As a second model we want to test whether population size has an impact. We use now a multivariate regression.
```{r Logistic_Regression, messages = FALSE}
mod2 <- lm(deaths_per_mill ~ cases_per_mill + Population, data = US_states_latest)
print(summary(mod2))
US_states_pred2 <- US_states_latest %>% mutate(pred = predict(mod2))
US_states_pred2 %>% ggplot() + geom_point(aes(x = cases_per_mill, y = deaths_per_mill),color="blue") + geom_point(aes(x = cases_per_mill, y = pred), color = "red")

```

## Conclusion
Deaths correlate with cases reported and the inclusion of the absolute population size improves the prediction slightly (higher adjusted R-squared). The reason therefore might be seen in the fact that locations with higher populations exhibit more contacts for each individual and hence a higher chance to get sick severely. 

## Bias
Be it for reported cases as well as for reported deaths one needs to be aware that there might be a bias on how data is collected and measured. How much tests per one million persons have been conducted, and how regular where those tests? Are deaths which where caused by COVID treated differently from deaths where the patient died from another disease while being COVID positive as well? I would expect that testing and reporting related to COVID differs from state to state, and also between cities and rural areas. Same is also true for the treatment of patients. How much medication was available and used? And what about the infrastructure? Availability of ICUs and their personal are definitely not equal across the entire US. Given that there are potentially many differences in reporting and also in the quality of treatment, the outcome of the regression model needs to be interpreted accordingly - meaning that we need to expect data points not to align perfectly to the regression line as we compare a little bit apples with oranges.

```{r session}
sessionInfo()
```